Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-24 src/tex/ms.tex
Finished editing sec 4

\section{Discussion} \label{sec:discussion}

The results in this work show a promising way to understand and improve the
accuracy of phenomenological waveform models by jointly optimizing all
coefficients at once. There are a number of caveats multiple ways to improve the
current work.

While this study focuses on the IMRPhenomD model because its availability as an
automatically differentiable waveform, other more sophisticate waveforms such as
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} and IMRPhenomXAS
\citep{pratten2020setting,pratten2021computationally} can potentially benefit
from recalibration even more, because of the larger number of calibration
parameters. % Extend this a bit?

The correlation we observe between the mismatches and the waveform's parameters
can also be used to design more physical ansatz. For example, we see the
recalibration process struggles to improve in the regime where $\chi_{1}$ is
positive and $\chi_{2}$ is negative, this means there is a dependency between
the two spins that is not captured by the ansatz. A systematic scheme that
incorporate this recalibration scheme with waveform design will be handy in
constructing new waveform models in the future. In making figure
\ref{fig:ps_q148_quadrant}, we split the parameter space into four regions
because of simplicity. These simple cuts demonstrate the accuracy of the
waveform can be further improved, but they are almost certainly not the optimal
way to incorporate the extra information we have about the waveform. A more
general way such as adding new functional dependency between the parameters
should be explored in the future.

We use mismatch to quantify the accuracy of the waveform model in this work. It
is a natural extension to investigate how would the recalibration affects
downstream analysis such as parameter estimation and population analysis.
Whether it is possible to reduce the systematic error in parameter estimation
using recalibrated waveforms is an interesting question we will investigate in
the sooner future. On top of this, recently relatively more differentiable
sampler such as \citep{Wong:2022xvh} has been introduced to the community, one
can also optimize the waveform directly using metrics from parameter estimation,
such as bias in an injection-recovery run. This will be an example of
more direct and efficient optimization of waveform models for downstream analysis.

@@

We have shown the result of recalibrating waveform coefficients. One thing to
note is that our recalibration procedure is not exactly the same as the original
calibration. For instance, we use a different set of NR waveforms, frequency
range, etc. Nonetheless, as the decrease in mismatch is rather significant, this
optimization procedure should be able to improve the accuracy of IMRPhenomD on a
similar scale regardless of the differences. Here, the result serves as a
demonstration of the general method used.

The results presented in Fig.~\ref{fig:q148_q1248_compare} demonstrate that
increasing the number of training waveforms used in waveform optimization yields
only a marginal increase in accuracy. Our analysis suggests that this marginal
improvement is a consequence of over-determination of the waveform coefficients.
Consequently, increasing the number of calibration NR waveforms is unlikely to
result in any significant improvement of the model's accuracy. These
observations suggest that the parameterized ansatz employed in our study may not
be suitable for certain regions in the parameter space, leading to low mismatches
for some waveforms while other waveforms remain at the high mismatch tail with
negligible changes. This highlights the constraints of the model's flexibility
that ultimately limit its performance.

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec.~\ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, which can be seen in 
Fig.~\ref{fig:ps_q148}. Notably, the ansatz performs better in the top-left
region than the bottom-right region. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec.~\ref{sec:result}. With separate optimizations for each
regions, Fig.~\ref{fig:ps_q148_quadrant} indicates that the top-left region's 
performance has improved, while the bottom-right region hardly improves. 
This observation suggests the ansatz is specific to certain regions of the 
parameter space, with a preference for BBH events lying above the diagonal, 
and it has limited enhancement for events lying below the diagonal.

The division of the parameter space into four regions was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically. Other than how to divide regions of interest, the 
choice of training waveforms also affects the final results greatly. Note that 
the choice of training waveforms listed in Tab.~\ref{tab:quadrants} were taken 
arbitrarily to test the effectiveness of separate fitting. Hence, if one takes 
a different set of training waveforms over the parameter space, the result might 
give additional features that can test and explain IMRPhenomD better.

@@

% Although our study primarily focused on the IMRPhenomD model, this simple yet
% versatile approach can be applied to other differentiable GW models, such as the
% IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} or IMRPhenomX \citep{pratten2020setting,pratten2021computationally} 
% models within the same family. By jointly optimizing a new set of coefficients,
% it is expected that both models can be enhanced since they share similar
% construction principles to the IMRPhenomD model. For instance, they also use PN
% approximants as part of the ansatz in the inspiral segment. It will be interesting 
% to recalibratethe IMRPhenomXAS model \citep{pratten2020setting}. Because it is
% parameterized by an additional anti-symmetric spin parameter, it is expected not
% to exhibit the degeneracy previously described. With the currently developing 
% {\jax} IMRPhenomXAS model in {\ripple}, A more detailed investigation
% may provide valuable insights into the systematics of the Phenom models.
% Furthermore, this approach is applicable to other GW model families, such as NR
% surrogate models or EOB models introduced in Sec. \ref{sec:intro}. Such an approach could 
% simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

>>>>

\section{Discussion and Conclusion} \label{sec:discussion}

The results in this work show a promising way to understand and improve the
accuracy of phenomenological waveform models by jointly optimizing all
coefficients at once. However, there are a number of caveats as well
as multiple ways to improve upon the current work.

While this study focuses on the IMRPhenomD model due to its availability as an
automatically differentiable waveform in \ripple, other more sophisticated waveforms such as
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} and IMRPhenomXAS
\citep{pratten2020setting,pratten2021computationally} can also potentially benefit
from recalibration. In fact, since these waveforms have a larger number of calibration
parameters and are fit to a larger number of NR simulations, it is possible that the 
improvement will be more dramatic than for IMRPhenomD. {\ripple} is being regularly
updated with new waveforms, and future work should carefully examine whether these more
modern waveforms can be improved. % Extend this a bit?

The correlation we observe between the mismatches and the waveform's parameters
can also be used to design more physical ansatz. For example, we see the
recalibration process struggles to improve in the regime where $\chi_{1}$ is
positive and $\chi_{2}$ is negative. This means there is a dependency between
the two spins that is not captured by the current ansatz. A systematic scheme that
incorporates our calibration method with waveform design will be handy in
constructing new waveform models in the future.

In making Fig.~\ref{fig:ps_q148_quadrant}, we split the parameter space into four regions purely for simplicity.
These simple cuts demonstrate the accuracy of the waveform can be further improved,
but they are almost certainly not the optimal way to incorporate the extra information
we have about the waveform. A more general way such as adding new functional dependency
between the parameters should be explored in the future.

In this work, we use the mismatch to quantify the accuracy of the waveform model. 
A natural extension of this work to investigate how the recalibration process affects
downstream analyses such as parameter estimation and population modeling.
Differentiable samplers such as~\citep{Wong:2022xvh}, which have been recently
introduced to the community, potentially allow one to optimize the waveform directly
using metrics from parameter estimation. For example, minimizing the bias in an injection-recovery run
could be used as a loss function. Overall, this approach could help reduce systematic waveform error
in parameter estimation simply through recalibrated waveforms. We plan to investigate these avenues in
the future.

@@

% As discussed in Sec.~\ref{sec:method} our recalibration procedure is not exactly the same as the original
% calibration. For instance, we use a different set of NR waveforms, frequency
% ranges, etc. Nonetheless, since the decrease in mismatch is rather significant, we believe that
% the high-dimensional optimization procedure is primarily responsible to the improved performance.

The results presented in Fig.~\ref{fig:q148_q1248_compare} indicate that
increasing the number of training NR waveforms used in the waveform optimization yields
only a marginal increase in accuracy. This observation suggests that the parameterized ansatz employed 
in IMRPhenomD struggles to capture the full complexity of the NR waveforms. 
However, as the accuracy requirements of waveform models increases with more sensitive detectors,
more NR waveforms will be required to train more flexible Phenom models. 
Accurate NR simulations must therefore be developed in parallel with waveform models
to ensure we meet future detector accuracy requirements.

Overall, the development of accurate waveform models is crucial for the success of GW astronomy.
In this work, we have explored how modern computational tools (automatic differentiation and gradient descent)
can help with this task. 
Our method is general and can be applied to any waveform model that is differentiable.
We therefore encourage the waveform development community to utlize these tools
and hope that this work can effetively contribute to the development of accurate waveform models.

% The reduced spin approximation is a major contributor to the inaccuracies
% observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
% parameter, $\chi_{\mathrm{PN}}$, as described in Sec.~\ref{sec:method}. The
% parameterization of BBH mergers using a single spin parameter results in a
% degeneracy within the parameter space. Specifically, black hole events with
% different spins may generate the same waveform due to identical values of
% $\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
% unequal spin events. This degeneracy produces straight lines in the parameter
% space with negative slopes that depend on the mass ratio, which can be seen in 
% Fig.~\ref{fig:ps_q148}. Notably, the ansatz performs better in the top-left
% region than the bottom-right region. In an attempt to address
% this issue, we partitioned the parameter space into four regions, as described
% in Sec.~\ref{sec:result}. With separate optimizations for each
% regions, Fig.~\ref{fig:ps_q148_quadrant} indicates that the top-left region's 
% performance has improved, while the bottom-right region hardly improves. 
% This observation suggests the ansatz is specific to certain regions of the 
% parameter space, with a preference for BBH events lying above the diagonal, 
% and it has limited enhancement for events lying below the diagonal.

\te{I suggest we end here}

The division of the parameter space into four regions was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically. Other than how to divide regions of interest, the 
choice of training waveforms also affects the final results greatly. Note that 
the choice of training waveforms listed in Tab.~\ref{tab:quadrants} were taken 
arbitrarily to test the effectiveness of separate fitting. Hence, if one takes 
a different set of training waveforms over the parameter space, the result might 
give additional features that can test and explain IMRPhenomD better. \te{I don't really
understand this paragraph. Should we remove it?}

@@

% Although our study primarily focused on the IMRPhenomD model, this simple yet
% versatile approach can be applied to other differentiable GW models, such as the
% IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} or IMRPhenomX \citep{pratten2020setting,pratten2021computationally} 
% models within the same family. By jointly optimizing a new set of coefficients,
% it is expected that both models can be enhanced since they share similar
% construction principles to the IMRPhenomD model. For instance, they also use PN
% approximants as part of the ansatz in the inspiral segment. It will be interesting 
% to recalibratethe IMRPhenomXAS model \citep{pratten2020setting}. Because it is
% parameterized by an additional anti-symmetric spin parameter, it is expected not
% to exhibit the degeneracy previously described. With the currently developing 
% {\jax} IMRPhenomXAS model in {\ripple}, A more detailed investigation
% may provide valuable insights into the systematics of the Phenom models.
% Furthermore, this approach is applicable to other GW model families, such as NR
% surrogate models or EOB models introduced in Sec. \ref{sec:intro}. Such an approach could 
% simplify NR waveform calibration procedures and lead to the improvement of 
% existing models.
